% sections/background.tex — Background
\section{Background}
\label{sec:background}

\subsection{The Actor Model}
\label{sec:bg:actor}

Hewitt, Bishop and Steiger~\citep{hewitt1973actors} introduced the
actor model in 1973 as a foundation for describing concurrent
computation in artificial intelligence systems.  The model rests on
three primitives: an actor processes messages from its mailbox one
at a time, may send messages to actors whose identifiers it knows,
and may create new actors.  No shared state, no locks.  Agha's
1986 dissertation~\citep{agha1986actors} gave the first formal
treatment, casting actor systems as labeled transition systems and
establishing the equivalences that justify mailbox semantics.
Subsequent work added typed actors~\citep{he2014typed},
capabilities~\citep{clebsch2015pony}, and supervision
trees~\citep{armstrong2007erlang}; the latter is the closest
ancestor of \system{}'s supervisor design.

Industrial adoption has been concentrated in distributed and
fault-tolerant settings.  Erlang/OTP~\citep{armstrong2007erlang} has
served the telecommunications industry since the late 1980s; the
Akka library~\citep{akka2010} brought the model to the JVM in 2010
and is widely deployed in reactive
microservices~\citep{vernon2015reactive}; Pony's deny-capability
type system~\citep{clebsch2015pony} makes data-race freedom a
compile-time property of actor message passing.  All of these
systems share an assumption \system{} does not inherit: that the
host operating system provides preemptive scheduling, heap-based
mailbox allocation, and cheap inter-actor synchronization through
mutexes or semaphores.  None of those assumptions transfer to the
GPU; the design contribution of \system{} is to rebuild the actor
abstraction on a substrate where they fail.

\subsection{GPU Programming Conventions}
\label{sec:bg:gpu}

NVIDIA's CUDA programming model~\citep{nickolls2008cuda} structures a
GPU computation as a \emph{kernel}—a function with implicit
parallelism—launched from the host across a two-dimensional grid of
thread blocks, each containing a configurable number of warps of 32
threads.  The host invokes a kernel with
\texttt{cuLaunchKernel}; the device runs it to completion across the
grid; the host then issues another kernel for the next phase of
computation.  State that needs to survive between kernels must be
explicitly placed in HBM (the device's high-bandwidth memory) by
the host before the first kernel and explicitly read back by the
host after the last.

Three architectural eras shape the model that we extend.

\paragraph{CUDA 1.0--3.x: kernels as functions}
The original CUDA~\citep{nickolls2008cuda} treats kernels as
stateless transformations: each launch begins with a fresh device
state, runs to completion, and exits.  Persistence across launches
is the host's responsibility, achieved through HBM allocation and
explicit \texttt{cudaMemcpy} calls.  Inter-block synchronization
within a single launch is impossible: blocks are scheduled to SMs
arbitrarily and may run, finish, and reuse SM resources before
sibling blocks even begin.  This era's only intra-launch
synchronization primitive is \texttt{\_\_syncthreads()} within a
single block.

\paragraph{CUDA 9.x: cooperative groups}
Cooperative Groups~\citep{nvidia2023cooperative} (CUDA~9, 2017)
introduced a typed hierarchy of synchronizable thread groups:
warps, blocks, multi-block groups, and—critically—the entire grid.
Grid-wide synchronization is enabled by launching a kernel with
\texttt{cuLaunchCooperativeKernel} (which guarantees that all blocks
of the grid are co-resident on the SMs simultaneously) and calling
\texttt{cooperative\_groups::this\_grid().sync()} from inside the
kernel.  This is the primitive that first makes a single kernel
launch capable of multi-stage computation without returning to the
host.  It is also the primitive that makes a persistent kernel
viable: a kernel can run a producer--consumer loop forever, with
\texttt{grid.sync()} as the barrier between iterations.

\paragraph{CUDA 12.x on Hopper: clusters, DSMEM, and async memory}
Hopper~(\citet{nvidia2022hopper}; CUDA 12.x) adds three primitives
on top of cooperative groups.  \emph{Thread block clusters} group up
to 16 blocks as a unit guaranteed to run on the same Graphics
Processing Cluster (GPC); \texttt{cluster.sync()} synchronizes within
that cluster at $\sim$0.6~$\mu$s ($\sim$2--3$\times$ faster than
grid-wide synchronization).  \emph{Distributed shared memory
(DSMEM)} lets each block in a cluster address the shared memory of
its sibling blocks via \texttt{cluster.map\_shared\_rank()}, with
$\sim$30 cycles of latency—an order of magnitude better than
HBM~\citep{nvidia2022hopper}.  \emph{Stream-ordered asynchronous
allocation} (\texttt{cuMemAllocAsync},
\texttt{cuMemFreeAsync}; CUDA~11.2 on Ampere, but stable on Hopper)
permits an actor to allocate and free memory inside its own
execution stream, replacing the global-synchronization penalty of
classical \texttt{cuMemAlloc} with a stream-local synchronization.
We measured an allocation-latency improvement of 116.9$\times$ on
H100 (878~ns versus 102.6~$\mu$s for 256-byte allocations);
\Cref{sec:evaluation} documents the measurement protocol.

\subsection{Persistent Kernels and Their Limits}
\label{sec:bg:persistent}

The persistent-kernel pattern was introduced informally by Aila and
Laine~\citep{aila2009raytrace} in the context of GPU ray traversal,
where the cost of relaunching a kernel for each ray dwarfed the
cost of the trace itself.  Their solution: launch a single kernel
that loops, pulling rays from a device-side work queue until the
queue is empty.  Gupta, Stuart, and Owens~\citep{gupta2012persistent}
gave the first systematic study of the pattern across a range of
GPGPU workloads, characterizing when persistent threads outperform
launch-per-iteration and when they don't.
Steinberger et~al.~\citep{steinberger2014whippletree} extended the
idea to dynamic workloads with the Whippletree scheduler, which
treats the persistent kernel as a runtime that schedules
heterogeneous tasks pulled from device-side queues.

In all three cases the kernel is a single producer--consumer loop:
one body of code, one queue, one termination condition.  The
abstraction is operationally similar to a thread pool with a single
shared work queue.  None of the three works treats individual
participants in the loop as actors with private state, lifecycle,
restart policy, or causal clock; the kernel is a scheduling trick
layered onto the kernel-launch model.  In \system{}'s language, all
three are special cases of a single actor population in which every
actor has the same behavior and the supervisor is trivial
(no restart, no per-actor snapshot, no per-actor isolation tag).

The gap our work fills is to lift the loop body to a population of
independently-supervised actors with the full complement of
lifecycle, supervision, causal time, and tenant isolation
primitives.  The gap is non-trivial because each of those
primitives requires device-side state that prior persistent-kernel
work did not need: per-actor snapshots in HBM, per-actor HLCs in
the state vector, per-actor audit tags routed through sub-brokers,
and a supervisor-managed restart counter with windowed escalation.
The lifecycle theorem of \Cref{thm:lifecycle-soundness} and the
restart-monotonicity theorem of \Cref{thm:hlc-restart-monotone}
have no analogue in the persistent-kernel literature because the
properties they assert do not exist in those systems.
