% sections/conclusion.tex — Conclusion
\section{Conclusion}
\label{sec:conclusion}

This paper defines and evaluates one plausible formal substrate for
long-lived, actor-style computation on GPUs: a small-step
operational semantics for the lifecycle of a device-resident actor,
a kernel-to-kernel messaging calculus with an explicit dead-letter
discipline, an HLC integration that survives supervisor-driven
restart, a routing-level cross-tenant isolation argument, and a
multi-GPU migration protocol.  Each is stated as a calculus,
analytically proved or sketched, and mechanized in \TLA{} under
bounded TLC model checking; the runtime realizing the calculus is
benchmarked on NVIDIA H100~NVL silicon.  To our knowledge this is
the first work that combines device-resident lifecycle,
supervision, causal clocks, tenant routing, and cross-GPU migration
under a single formal model.

The model can be read as relating three design patterns that are
not usually treated together in the GPU systems literature:
persistent kernels (the lifecycle without supervision or causal
time), device-side work queues (the K2K calculus without audit
boundaries), and supervisor trees (the supervision discipline
without a formal substrate).  Formalizing the substrate makes
each of those patterns expressible as a special case within a
common framework rather than as a separate engineering device.

The technical contributions—the lifecycle calculus
(\Cref{sec:model}), the K2K messaging discipline
(\Cref{sec:k2k}), the HLC integration with restart
(\Cref{sec:supervision}), and the cross-tenant isolation theorem
(\Cref{sec:tenancy})—were chosen to be the smallest set of
properties that an audit-grade or compliance-grade GPU runtime
must demonstrate.  Each is mechanized under bounded configurations;
each is mapped to a Rust module in the open-source
\texttt{ringkernel-core} runtime; each is cross-referenced against
H100 measurements that ship with the distribution under
\texttt{docs/benchmarks/}.  The verification pass caught a real
spec-level bug in the migration checksum invariant
(\Cref{sec:disc:tlc-bug}), which is the kind of engineering outcome
we think formal methods should be held to.

The empirical results---the 8{,}698$\times$ command-submission
latency figure, 5.10~M~msg/s sustained throughput at CV 0.66\%,
0.628~$\mu$s cluster-level synchronization, 116.9$\times$ async
memory allocation improvement---do not by themselves justify the
actor model on the GPU; they support the narrower proposition that,
having chosen an actor-style substrate for a GPU runtime, the
Hopper-era primitives sit within the cost envelope that actor
systems typically require of their substrates.  The 4.2$\times$
slowdown of the block-actor WaveSim3D relative to a hand-tuned
stencil records the other side of the ledger: there is a
measurable abstraction cost, and it should be budgeted for in
application choice.  The aim of the paper is not to claim that the
actor model is always the appropriate GPU abstraction; it is to
offer a formal and empirical basis on which to decide.

\paragraph{Open artifacts}
All artifacts referenced in this paper are available in the
\system{} open-source distribution under MIT license:

\begin{itemize}
\item \texttt{docs/verification/} — the \TLA{} specifications,
\texttt{tlc.sh} runner, and per-spec configuration files.
\item \texttt{crates/ringkernel-core/}, \texttt{ringkernel-cuda/},
\texttt{ringkernel-cuda-codegen/} — the runtime source.
\item \texttt{docs/benchmarks/} — the methodology document, the
academic-proof report, and the H100/B200 baseline characterization.
\item \texttt{target/criterion/} (after running the benchmark
suite) — raw data, including per-trial timings and confidence
intervals.
\item \texttt{docs/paper/} — the source of this paper, including
the build automation in \texttt{Makefile}.
\end{itemize}

The repository is at the URL stated in the paper metadata; pull
requests, formal-model extensions, and adversary-aware additions
are welcome.
