% sections/discussion.tex — Discussion
\section{Discussion}
\label{sec:discussion}

\subsection{What the Model Does Not Cover}
\label{sec:disc:limits}

We collect, in one place, the ways in which the formal model
abstracts over the runtime.  Each is a candidate for future work
and a limitation of the current paper's claims.

\paragraph{Warp divergence}
The model treats a \textsc{Process} step as atomic; in the runtime,
a \textsc{Process} step is executed by a warp of 32 threads in SIMT
fashion, and divergent control flow within the warp serializes
through predicated execution.  The cost model the runtime presents
to applications is conservative (assume the worst-case branch);
the formal calculus is silent on divergence.  A finer-grained
calculus that models per-thread state and predicated execution
would let one prove tighter bounds for divergent actor bodies but
would substantially complicate every theorem.  We chose the
coarser abstraction.

\paragraph{Cooperative-group fault semantics}
Cooperative kernel launches assume that all blocks of the grid are
co-resident on the SMs; if an SM faults (e.g., ECC uncorrected
error), the entire kernel is torn down and the supervisor must
restart all actors of the failed grid.  The formal model treats
this as ``every actor's life'' transitioning to \Failed{} via a
single supervisor-issued \textsc{Fail} step; the actual
device-level fault recovery is not modeled.  A more complete
model would distinguish recoverable per-block faults from
unrecoverable per-grid faults; we leave this to future work.

\paragraph{Migration: proof sketch plus bounded mechanization, not
full analytical proof}
The migration protocol of \Cref{sec:migration} is stated as a phase
machine, and \Cref{thm:migration-safety} is given with a proof
sketch and discharged by bounded TLC in
\texttt{docs/verification/migration.tla} and
\texttt{multi\_gpu\_k2k.tla}.  We do not give a full analytical
proof in this paper.  Two complications are left to a companion
treatment: the cross-device HLC merge argument of
\Cref{sec:migration:hlc-cross} under non-negligible host
wall-clock skew, and DSMEM re-establishment on the destination
(DSMEM is per-device, so the destination device must rebuild the
cluster topology before the actor resumes).  The bounded
mechanization rules out counterexamples within its configuration
but does not discharge the theorem at arbitrary scale.

\paragraph{Adversary model for tenant isolation}
The cross-tenant-isolation theorem
(\Cref{thm:cross-tenant-isolation}) is proved against the calculus's
own routing rule.  It is not proved against an adversary who can
subvert the runtime—e.g., by forging an audit tag, by writing
directly into another tenant's HBM region, or by exploiting a
device-level memory-protection failure.  Such an adversary is
outside the scope of the formal model; defending against it is a
question of runtime hardening (we discussed defense in depth in
\Cref{sec:tenancy:limits}).  An adversary-aware version of the
isolation theorem would require modeling the device's memory
protection guarantees, which depend on hardware features (e.g.,
Multi-Instance GPU, Confidential Computing for Hopper) we have
not yet incorporated.

\subsection{What the Verification Pass Caught}
\label{sec:disc:tlc-bug}

The model-checking pass against the six \TLA{} specifications
(\Cref{sec:eval:tlc}) is normally a confirmation step — it checks
that the spec the paper describes still satisfies the invariants
the paper claims.  In the v1.1 verification pass it caught a real
specification-level bug.

The migration spec (\texttt{migration.tla}) declares a
\texttt{ChecksumMatch} invariant: at the moment of
\textsc{SwapRouting}, the source's captured checksum equals the
target's restored checksum (\Cref{thm:migration-safety}, clause~4).
The original spec computed both checksums from the live
\texttt{src\_state} variable.  But \texttt{AcceptDuringQuiesce}
(the action that handles late messages arriving during
\state{Quiesced}, \state{Transferring}, and \state{Swapping})
mutates \texttt{src\_state} while the migration is in progress.
A trace exists in which a late message lands in
\texttt{src\_state} between \texttt{CaptureState} and
\texttt{Transfer}; the source's checksum is recomputed before the
swap and now disagrees with the target's restored checksum.  TLC
reported the counterexample after $\sim$2 minutes of search.

The Rust implementation was already correct: it captures the
staging buffer as a single immutable byte slice at
\textsc{CaptureState} and both sides compute CRC32 over those
bytes.  The spec's bug was that it did not model the immutability
of the captured snapshot.  We patched the spec to introduce an
explicit \texttt{captured\_state} variable, frozen at
\texttt{CaptureState}, used by both \texttt{Transfer} and the
checksum check.  After the patch the invariant holds under the
full bound.

The episode is of the kind that formal methods are often cited as
being useful for: a divergence between the spec's idealization of
an implementation detail and the implementation itself.  In this
case the implementation happened to be the more careful party; in
another project the asymmetry could go the other way.  The
verification pass is what surfaces the disagreement.

\subsection{Open Questions}
\label{sec:disc:open}

\paragraph{Scaling the verification}
The bounded \TLA{} verification covers parameter ranges of 2--3
actors per spec, 4--10 messages per spec, queue capacity 2--4.  This
is sufficient to expose state-machine bugs (the kind that show up at
small configurations or do not show up at any) but does not cover
production scales (thousands of actors, millions of messages).
Scaling the verification will likely require symbolic abstraction
techniques or compositional reasoning; a TLAPS-style proof of a
scale-independent invariant, separately validated by TLC at small
scales, is one direction.

\paragraph{Heterogeneous backends}
The formal model is GPU-agnostic in its statement (no rule
references CUDA-specific primitives) but GPU-specific in its
implementation: AMD HIP and Intel oneAPI offer analogous
primitives, but the cost models differ enough that the same
calculus on a different backend would produce a different
performance profile.  A unified backend abstraction in
\texttt{ringkernel-core} that maps the same high-level rules onto
CUDA, HIP, and oneAPI realizations is the largest engineering
direction we see.

\paragraph{NSAI (neuro-symbolic AI) workloads}
The motivating workload for \system{} is neuro-symbolic
reasoning: a population of actors representing logical rules, a
neural-network actor representing perception, a supervisor
actor coordinating restart on rule-engine timeouts.  We expect to
report on this workload in a follow-up paper.  The formal model
of this paper is the substrate; the NSAI application is the test
of whether the substrate's primitives compose into something
domain-experts can reason about productively.

\paragraph{Comparison against alternative GPU programming abstractions}
We focused this paper on the formal model and validated the
empirical numbers against the formal claims.  A comprehensive
performance study against Triton, JAX, and CUDA Graphs across
diverse workloads is out of scope here; we expect specific
workloads will favor each of those abstractions over \system{},
and equally specific workloads will favor \system{} over each of
them (in particular, workloads with long-lived per-actor state and
fine-grained inter-actor messaging).  Sketching the trade-off
landscape is future work.

\subsection{Future Hardware}
\label{sec:disc:hw}

The formal model is parameterized by a small set of hardware
primitives: cluster size, DSMEM scope and latency, cooperative
launch semantics, atomic-allocation cost, and the per-device
fault-detection surface.  Two NVIDIA architectures beyond Hopper
are relevant: Blackwell (B100/B200, GB200~NVL72, shipping; features
documented in NVIDIA's public tuning and whitepaper materials) and
Rubin (R100/R200, NVL144, in production Q1~2026; figures are
NVIDIA's announced specifications and subject to change before
silicon is publicly evaluable).  We distinguish between the two
throughout this subsection.  We split the extensions into
\textbf{documented} (grounded in current vendor documentation and
published microbenchmarks) and \textbf{speculative} (contingent on
announced specifications whose details have not been publicly
confirmed).  We catalog each as a starting point for follow-on
work; we do not predict performance.

\paragraph{Documented extensions (Blackwell)}
The items in this block rest on current NVIDIA documentation and
published microbenchmarking results for Blackwell silicon.

\paragraph{Cluster size and per-cluster supervision}
Blackwell preserves Hopper's cluster geometry: up to 8 blocks
portable, 16 blocks
nonportable~\citep{nvidia2024blackwell-tuning,colfax2024clusters}.
Cluster size therefore is not a Blackwell-specific extension.  The
formal model's per-supervisor child count
(\Cref{sec:supervision:policies}) is an implementation parameter
rather than a soundness parameter, so it does not affect
\Cref{thm:lifecycle-soundness} regardless.  The relevant new
opportunity is to prove a \emph{cluster-locality} bound on a
fault's blast radius: a supervisor whose children all reside in
the same nonportable-size-16 cluster has cross-cluster K2K traffic
only when escalating to its own supervisor.  This bound is
expressible on Hopper too, but Blackwell's larger
shared-memory-per-block (up to 227~KiB on
B200~\citep{nvidia2024blackwell-tuning}) makes 16-actor clusters
practical for state-heavy supervisor workloads.  Whether Rubin
changes cluster geometry is not publicly disclosed at the time of
writing (NVIDIA's Rubin platform documentation describes per-die
SM count and NVLink-6 bandwidth but not cluster
geometry~\citep{nvidia2025rubin}).

\paragraph{Subset-synchronization barriers}
Blackwell's \texttt{tcgen05.commit} bitmask synchronization
primitive~\citep{colfax2024clusters} permits a cluster-scoped
barrier on a strict subset of the cluster's blocks rather than the
entire cluster.  In K2K terms, this lets the
\textsc{Dsmem} tier of \Cref{sec:k2k:tiers} synchronize only the
blocks involved in a given conversation, rather than draining the
whole cluster's tick.  A further theorem becomes stateable:
\emph{per-conversation liveness}—if blocks $B_1, B_2$ exchange a
message, the conversation makes progress without quiescing the
cluster's other blocks.  The current K2K calculus does not state
this; doing so would require a refinement of the tick semantics
to expose the subset structure.  We regard this as one of the
natural candidates for a Blackwell-targeted refinement of the
model.

\paragraph{Confidential computing for tenant isolation}
Blackwell is the first NVIDIA GPU with TEE-I/O capability
(hardware-enforced encrypted I/O paths) and Multi-Instance GPU /
Confidential Computing
composability~\citep{nvidia2024secureai}.  Rubin
extends this to NVLink-C2C (CPU--GPU) and NVLink (GPU--GPU)
encryption~\citep{nvidia2025rubin}.  The cross-tenant isolation
theorem (\Cref{thm:cross-tenant-isolation}) is currently a
structural argument over the runtime's routing function, with the
limitation noted in \Cref{sec:tenancy:limits} that the model does
not formalize the adversary's hardware capabilities.  A
Blackwell-or-later target lets the theorem be stated against an
adversary with the standard hardware-attacker model: the
encrypted I/O paths reduce the adversary's surface to one that the
TEE specification already characterizes, and the isolation theorem
becomes a corollary of the TEE's confidentiality guarantee
composed with the runtime's routing function.  This would close
the gap between the structural argument and the cryptographic
argument.

\paragraph{Speculative extensions (Rubin and rack-scale follow-on)}
The remaining items rest on NVIDIA's announced Rubin
specifications; cluster geometry, DSMEM scope, and
\texttt{cluster.sync()} latency on Rubin have not been publicly
disclosed at the time of writing.  Each should be re-examined once
silicon is available.

\paragraph{Self-healing actor systems (Rubin)}
Rubin's second-generation RAS engine includes in-field SRAM repair
and zero-downtime self-testing~\citep{nvidia2025rubin}, replacing
Blackwell's detection-and-localization capability with active
remediation.  The actor model gains the ability to migrate
actors off a degrading SM \emph{before} the SM fails, rather than
restart actors after a fault has already occurred.  A new
predictive-restart rule could be added to
\Cref{sec:supervision:policies}; the corresponding safety
property would be that the new instance starts before the old
instance's HLC is observable to be ``stuck.''  Stating and
mechanizing this requires modeling the supervisor's view of
device-health telemetry, which is currently outside the model.

\paragraph{Rack-scale collapses the K2K tier hierarchy}
GB200~NVL72 (Blackwell) and the Rubin~NVL144 follow-on present 72
or 144 GPUs as a single all-to-all NVLink fabric (1.8~TB/s and
3.6~TB/s per GPU, respectively~\citep{nvidia2024gb200,%
nvidia2025rubin}).  At this scale, the cost gap between intra-cluster
K2K (\textsc{Dsmem}) and inter-GPU K2K (NVLink P2P, per
\Cref{sec:migration:cross-gpu}) narrows to a single order of
magnitude rather than two; the fallback-to-host path of
\Cref{thm:fast-path-selection} does not occur for any pair of
GPUs in the rack.  The formal model's three-tier K2K hierarchy can
be collapsed to two tiers in this regime (intra-cluster and
intra-rack), with a new \emph{rack-locality} theorem stating that
no cross-rack message exists in a single-rack deployment.  The
operational reading is that an entire rack acts as a logical GPU
to the actor system; the supervisor tree of
\Cref{fig:supervisor-tree} extends naturally without changing the
restart-policy semantics.

\paragraph{Hardware decompression as a system actor}
Blackwell's Decompression Engine is a fixed-function block on the
copy engine, executing LZ4/Snappy/Deflate at $\sim$800~GB/s in
parallel with SM kernels~\citep{nvidia2024decomp}.  Within our
formal model, this maps cleanly to a \emph{system actor}: an actor
whose \(\delta\) function is implemented in hardware rather than
in CUDA, with bounded per-tick throughput and no SM-cycle
consumption.  Adding a new actor class for hardware-implemented
\(\delta\) functions—keeping the lifecycle, K2K, and supervisor
machinery, but bypassing the persistent-kernel substrate for the
\textsc{Process} step—is a small extension that opens the model to
a class of accelerator primitives that are otherwise difficult to
expose in actor-style code.

\paragraph{Tensor Memory: a constraint, not an extension}
Blackwell's per-SM 256~KiB Tensor Memory (TMEM) is reachable only
via \texttt{tcgen05.*} instructions, not from ordinary
SMEM/DSMEM addresses~\citep{microbench-blackwell}.  Actor state
must therefore continue to live in HBM, SMEM, DSMEM, or
registers; TMEM is a substrate for matrix operations that the
\(\delta\) function may invoke but not a substrate for actor
state.  This is worth saying explicitly because a reader might
otherwise expect a 256~KiB on-chip extension of \(\mathit{snap}\)
to be available; it is not.

\paragraph{Caveats on the road map}
We treat the speculative items above as research \emph{directions}
contingent on the as-yet-unpublished Rubin parameters; a
formal-model update for any specific Rubin extension would follow a
hardware-availability checkpoint similar to the one the H100~NVL
evaluation of \Cref{sec:evaluation} represents for Hopper.  The
documented Blackwell items are better supported but still require
that we validate the hardware behavior we model—microbenchmarks
first, extensions to the calculus second—before the corresponding
theorems can be claimed.

\subsection{A Broader Framing}
\label{sec:disc:big}

The framing we would suggest, and would be happy to have disputed,
is that recent GPU architectures have begun to expose primitives
that systems work can plausibly target, not just numerics.  The
coprocessor framing---the host writes a batch of data, calls a
function, reads the results back---has served GPU programming for
most of its history.  The Hopper-era additions of clusters, DSMEM,
asynchronous allocation, and cooperative launches are performance
features in the first instance, but they also make it possible to
treat the GPU as a long-running computational fabric in ways that
earlier architectures did not readily support.  This paper takes
one step in that direction by treating the fabric as a target for
the actor model.  Whether analogues of other systems abstractions
(databases, queues, schedulers, supervisors of other kinds) will
sit as naturally on the same fabric, and whether the
formal-model-then-mechanization methodology we apply here will
carry over, is an open question.
