% sections/evaluation.tex — Evaluation
\section{Evaluation}
\label{sec:evaluation}

The evaluation is structured around the formal model: each
benchmark either exercises a specific operational primitive on
which the model depends, or measures a system-level property the
model anticipates.  We do not claim a comprehensive comparison
against all GPU programming abstractions; we benchmark against the
host-driven kernel-launch baseline and against CUDA Graphs (the
strongest publicly-documented host-driven alternative we are aware
of), and report the numbers that the model's quantitative claims
rest on.

All measurements were collected on 2026-04-16 on a single H100~NVL
under the conditions described in
\Cref{sec:eval:methodology}; raw data is available in
\texttt{target/criterion/} in the open-source distribution, and the
methodology document is \texttt{docs/benchmarks/METHODOLOGY.md}.

\subsection{Methodology}
\label{sec:eval:methodology}

We follow the statistical protocol of \citet{georges2007rigorous}
on confidence-interval reporting and between-condition significance
testing, augmented for parallel systems with the reporting
guidelines of \citet{hoefler2015scientific} on clock locking,
exclusive-access configuration, and coefficient-of-variation
discipline.  Concretely:

\begin{itemize}
\item GPU clocks are locked at 1{,}785~MHz via
\texttt{nvidia-smi -lgc}; persistence mode is enabled
(\texttt{nvidia-smi -pm 1}); compute mode is set to
\texttt{EXCLUSIVE\_PROCESS} to preclude background contention.

\item Each measurement is replicated across 20 independent trials,
each consisting of 1{,}000 inner iterations; the per-iteration
timing is reported with a 95\% confidence interval.

\item Between-condition comparisons (e.g., persistent actor vs.\
\texttt{cuLaunchKernel}) report Cohen's~\(d\) and a Welch's
\(t\)-test \(p\)-value; the difference is considered significant if
\(p < 10^{-3}\) and \(|d| > 2\) (very large effect).

\item All measurements are taken on Azure
\texttt{Standard\_NC80adis\_H100\_v5} (AMD EPYC 9V84 80-vCPU,
314~GiB DDR5, 2\(\times\) H100~NVL with 95{,}830~MiB HBM3 each
and ECC enabled, NV12 NVLink topology between GPU0 and GPU1),
CUDA Toolkit
12.8 (V12.8.93), driver 595.58.03.

\item ECC error counts are sampled before and after each trial;
trials in which any ECC corrected error appears are discarded
(zero discards across the reported runs).
\end{itemize}

\subsection{Command Submission Latency}
\label{sec:eval:injection}

The first claim is about the \emph{control path}: the host-side
cost of submitting one additional unit of work to the device
through three different work-submission mechanisms.  This is a
latency measurement for the control-path operation—mapped-memory
store versus \texttt{cuLaunchKernel} versus CUDA Graph replay—not a
runtime-against-runtime comparison of execution speed.  The three
mechanisms dispatch work on different units of orchestration (a
command slot in a persistent-kernel's control block versus a fresh
kernel launch versus a graph-replay submission), so the comparison
is fair \emph{for the question posed} (per-command host overhead)
and informative about when each mechanism dominates cost, but it
does not establish that one runtime is intrinsically faster than
another for a given workload.  We compare the three mechanisms in
\Cref{tab:injection}.

\begin{table}[htbp]
\centering
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lrrrr}
\toprule
Mechanism & Amort.\ per-cmd ($\mu$s) & Burst total ($\mu$s) & 95\% CI ($\mu$s) & vs.\ baseline \\
\midrule
\texttt{cuLaunchKernel} (baseline)   & 1.583   & 1{,}583.1 & $\pm$6.0  & 1.0$\times$ \\
CUDA Graph replay                    & 0.547   &    546.9  & $\pm$12.3 & 2.9$\times$ \\
Persistent actor (mapped-mem store)  & 0.00018 &      0.182 & $\pm$0.01 & \textbf{8{,}698$\times$} \\
\bottomrule
\end{tabular}
\caption{Host-side submission cost for a 1{,}000-command burst,
across three work-submission mechanisms: classical host launch
(one \texttt{cuLaunchKernel} per command, with
\texttt{cuStreamSynchronize} at burst end); CUDA Graph (1{,}000
launches captured, then one \texttt{cuGraphLaunch} replay); and
persistent actor (1{,}000 \texttt{ptr::write\_volatile} writes to
mapped memory, followed by one x86 \texttt{mfence}).  20 trials,
95\% CI from trial-level variance.  ``Burst total'' is the
wall-clock for the full 1{,}000-command burst; ``per-cmd
amortized'' is burst~/~1000 and should be read as the amortized
per-command cost through the CPU's write-combining buffer, not
as the one-shot propagation latency of a single write (which
includes PCIe crossing and is measured separately; order of
magnitude \(\sim\)100--300\,ns with a flush on H100~NVL).  The
8{,}698$\times$ ratio compares burst totals and means ``the host
spends 8{,}698$\times$ less time submitting this burst on this
path'', not ``persistent actors execute kernels 8{,}698$\times$
faster''.}
\label{tab:injection}
\end{table}

The Welch's \(t\)-test on the persistent-actor versus
\texttt{cuLaunchKernel} comparison gives \(p < 10^{-20}\) with
Cohen's \(d \gg 2.0\); the persistent-actor versus CUDA-Graph
comparison is similarly significant (\(p < 10^{-20}\), \(d \gg
2.0\)).  The 8{,}698$\times$ figure is the largest single
control-path gap we report; within the burst comparison, it is
what lets the persistent-kernel substrate absorb command bursts
without the host becoming the bottleneck.  It is not a claim
about execution throughput once a command has been accepted---%
that is what \Cref{sec:eval:throughput} measures.  The reported
amortized per-command cost assumes write-combining semantics; a
single isolated mapped-memory write with an explicit flush is
slower (a few hundred nanoseconds), and we report K2K
message-passing round-trip latency directly in
\Cref{sec:eval:throughput} rather than inferring it from this
table.

The interpretation, in terms of the formal model, is that the
\textsc{Process} step's host-side setup cost (the host's
contribution to ``ready the actor for one tick'') is negligible in
the persistent-kernel model and substantial in either of the
host-orchestrated alternatives.  Workloads in which the actor's
work per tick is small (telemetry aggregation, sensor fusion,
streaming graph updates) can amortize the host's submission cost
only in the persistent-kernel design.

\subsection{K2K Throughput and Stability}
\label{sec:eval:throughput}

The second claim is that the K2K calculus's lock-free queues sustain
high throughput at low latency, with stability sufficient for
operational use.  We measured per-message latency at four payload
sizes (\Cref{tab:k2k-latency}) and a 60-second sustained run at the
1~KiB payload size to characterize stability
(\Cref{tab:k2k-sustained}).

\begin{table}[htbp]
\centering
\begin{tabular}{lrrr}
\toprule
Payload & Latency (ns) & 95\% CI (ns) & Throughput (M~msg/s) \\
\midrule
64 B   & 72.28  & [72.26, 72.32] & 13.83 \\
256 B  & 74.67  & [74.65, 74.70] & 13.39 \\
1 KiB  & 82.64  & [82.61, 82.68] & 12.10 \\
4 KiB  & 177.50 & [177.47, 177.53] & 5.63 \\
\bottomrule
\end{tabular}
\caption{K2K per-message latency by payload size, 1{,}000 messages
per trial, 20 trials, in-block (\textsc{Smem}) tier.}
\label{tab:k2k-latency}
\end{table}

\begin{table}[htbp]
\centering
\begin{tabular}{lr}
\toprule
Metric & Value \\
\midrule
Duration per trial & 60.0 s \\
Trials & 4 \\
Mean throughput & \textbf{5.10~M~msg/s} \\
Coefficient of variation (across all 48 windows) & \textbf{0.66\%} \\
p50 latency & 90 ns \\
p95 latency & 100 ns \\
p99 latency & 110 ns \\
Throughput drift (first vs.\ last window per trial) & $<$2\% \\
Memory growth & 0 MiB \\
ECC errors (pre $\to$ post per GPU) & 0 / 0 \\
\bottomrule
\end{tabular}
\caption{K2K sustained run, four 60-second trials at 1~KiB payload
on 2$\times$ H100~NVL.  Per-window samples taken every 5 seconds (12
windows per trial); the CV of 0.66\% across the resulting
48-window pool sits well below the conventional ``extremely
stable'' threshold of 1\%.  Per-window p99 latency is flat at
110~ns across all 48 windows — no thermal throttle, no drift.}
\label{tab:k2k-sustained}
\end{table}

The throughput in \Cref{tab:k2k-sustained} is lower than the
inverse of the 90~ns~p50 latency would suggest: the gap is the
device-side BSP barrier between phases of the persistent kernel
(per \Cref{sec:implementation:kernel}).  Each iteration of the
kernel processes one message per actor; the throughput is the
product of actor count and tick rate.  At 5.10~M~msg/s with a
4-block, 256-thread cluster, this is roughly 5.0~K
messages/actor/second, sustained for 60 seconds without drift.

\paragraph{Per-tier round-trip latency}
The K2K channel hierarchy of \Cref{sec:k2k:tiers} predicts a
clean monotonic ordering \textsc{Smem}~$<$~\textsc{Dsmem}~$<$~\textsc{Hbm}.
\Cref{tab:k2k-tier} measures all three tiers at the same four
payload sizes via a host-observed kernel launch + synchronization
round-trip (the relevant operational cost for the actor system,
which always reaches each tier through a launch).  The per-message
\emph{device-cycle} latency is sub-microsecond on every
tier~\citep{nvidia2022hopper}; what
\Cref{tab:k2k-tier} reports is the host-side wall-clock that the
actor runtime actually pays.

\begin{table}[htbp]
\centering
\begin{tabular}{lrrrr}
\toprule
Payload & SMEM (ns) & DSMEM (ns) & HBM (ns) & HBM/SMEM \\
\midrule
64 B   & 6{,}793 $\pm$ 41 &  9{,}012 $\pm$ 40 & 10{,}460 $\pm$ 39 & 1.54$\times$ \\
256 B  & 6{,}702 $\pm$ 31 & 10{,}598 $\pm$ 34 & 12{,}066 $\pm$ 33 & 1.80$\times$ \\
1 KiB  & 6{,}722 $\pm$ 34 & 14{,}059 $\pm$ 34 & 15{,}811 $\pm$ 145 & 2.35$\times$ \\
4 KiB  & 6{,}749 $\pm$ 43 & 15{,}289 $\pm$ 106 & 18{,}237 $\pm$ 30 & 2.70$\times$ \\
\bottomrule
\end{tabular}
\caption{K2K tier-by-tier host-observed wall-clock, 1{,}000 trials
per cell, 200 warmup iterations.  SMEM is payload-independent
(state lives in each block's shared memory; the launch cost
dominates).  DSMEM scales with payload as data flows through the
DSMEM addressing path.  HBM adds the global-memory round-trip on
top of DSMEM, growing the HBM/SMEM ratio from 1.54$\times$ at 64~B
to 2.70$\times$ at 4~KiB — clean monotonic hierarchy across every
payload size, validating the calculus in \Cref{sec:k2k:tiers}.}
\label{tab:k2k-tier}
\end{table}

\subsection{Cluster-Level Synchronization}
\label{sec:eval:cluster}

The third claim is that Hopper's cluster-scoped synchronization is
substantially cheaper than grid-wide synchronization, and that
this difference is exploitable by the K2K calculus's tier choice
(\Cref{sec:k2k:tiers}).  \Cref{tab:cluster-sync} compares
\texttt{cluster.sync()} and \texttt{grid.sync()} under identical
conditions.

\begin{table}[htbp]
\centering
\begin{tabular}{lrrrr}
\toprule
Method & Per-sync ($\mu$s) & Total ($\mu$s) & 95\% CI ($\mu$s) & Std Dev ($\mu$s) \\
\midrule
\texttt{cluster.sync()} & \textbf{0.628} & 628.2 & $\pm$1.4 & 3.1 \\
\texttt{grid.sync()}    & 1.875          & 1{,}874.9 & $\pm$3.5 & 8.0 \\
\bottomrule
\end{tabular}
\caption{Synchronization latency, 1{,}000 syncs per trial, 20
trials, 4-block $\times$ 256-thread workload.}
\label{tab:cluster-sync}
\end{table}

The 2.98$\times$ speedup of \texttt{cluster.sync()} over
\texttt{grid.sync()} (\(p < 10^{-20}\), Cohen's \(d \gg 2.0\))
is consistent with the architecture: cluster-scoped barriers
synchronize only blocks within a single GPC, while grid-scoped
barriers synchronize across all GPCs and incur the GPC-to-GPC
communication cost.  In K2K terms, this means the
\textsc{Smem} and \textsc{Dsmem} tiers can iterate at
0.628~$\mu$s/tick while the \textsc{Hbm} tier is bounded by the
1.875~$\mu$s/tick of \texttt{grid.sync()}.  The K2K tier choice is
not only a memory-tier choice; it is also a synchronization-tier
choice.

\subsection{Asynchronous Memory Allocation}
\label{sec:eval:async-mem}

The fourth claim is that stream-ordered asynchronous allocation
(\Cref{sec:bg:gpu}, third paragraph) eliminates the global driver
lock that dominates the cost of classical \texttt{cuMemAlloc}.
\Cref{tab:async-alloc} reports the comparison at a 256-byte
allocation size.

\begin{table}[htbp]
\centering
\begin{tabular}{lrr}
\toprule
Method & Per-alloc (ns) & Speedup \\
\midrule
\texttt{cuMemAlloc} (synchronous) & 102{,}576 & 1.0$\times$ \\
\texttt{cuMemAllocAsync} (stream-ordered) & \textbf{878} & \textbf{116.9$\times$} \\
\bottomrule
\end{tabular}
\caption{Allocation latency, 256-byte allocations, 1{,}000
allocations per trial, 20 trials.}
\label{tab:async-alloc}
\end{table}

The 116.9$\times$ improvement is what permits the
``near-free actor allocation in the actor's own stream'' assumption
of \Cref{sec:implementation:driver}: at 102~$\mu$s per allocation,
spawning a thousand actors would consume 100~ms of host time and
serialize against every other CUDA stream; at 878~ns, the same
thousand spawns consume under 1~ms and overlap with other stream
work.

\subsection{Lifecycle Rule Overhead}
\label{sec:eval:lifecycle}

The lifecycle calculus of \Cref{sec:model:lifecycle} prescribes one
SOS rule per supervisor-issued event (\textsc{Spawn}, \textsc{Activate},
\textsc{Quiesce}, \textsc{Terminate}, \textsc{Restart}).  The
host-side cost of each rule is the time from the supervisor writing
the command into mapped memory to reading back the device-issued
acknowledgement.

\begin{table}[htbp]
\centering
\begin{tabular}{lrrr}
\toprule
Rule & Mean (ns) & Std dev & p99 (ns) \\
\midrule
\textsc{Spawn}     & 23.1 & 4.7 & 30 \\
\textsc{Activate}  & 23.4 & 4.7 & 30 \\
\textsc{Quiesce}   & 23.8 & 4.8 & 30 \\
\textsc{Terminate} & 23.7 & 4.8 & 30 \\
\textsc{Restart}   & 23.1 & 4.6 & 30 \\
\bottomrule
\end{tabular}
\caption{Lifecycle rule wall-clock, 500 trials per rule, 50
warmup iterations.  All five rules cluster around 23~ns mean /
30~ns p99 — flat within measurement noise.  This is the cost
of issuing a single SOS rule from the host; the device-side
processing overlaps with the next supervisor action through the
mapped-memory ring.}
\label{tab:lifecycle}
\end{table}

The 23-ns mean is two orders of magnitude below the
\emph{sub-microsecond} ceiling we set in \Cref{sec:disc:limits}; in
operational terms, a supervisor can issue 40~M~rule events/second
to the device without saturating the mapped-memory channel.  The
cost is the same for every rule because the device-side dispatch
is a single switch on the command opcode — no per-rule state
machinery on the issue path.

\subsection{Multi-GPU NVLink Bandwidth}
\label{sec:eval:mgpu-bw}

For the cross-GPU K2K calculus of \Cref{sec:migration:cross-gpu},
the principal question is whether the runtime can saturate NVLink
P2P bandwidth at realistic transfer sizes.  We measured back-to-back
\texttt{cuMemcpyPeerAsync} bursts (256 transfers per measurement,
32 warmup iterations, single stream) on a 2$\times$ H100~NVL
system with the 12-link NVLink4 bundle: each link runs at 25~GB/s
per direction, so the bundle's theoretical peak is 300~GB/s
unidirectional (equivalently 600~GB/s bidirectional).  The
\texttt{cuMemcpyPeerAsync} path is one-way, so we report
utilisation against the 300~GB/s unidirectional peak.

\begin{table}[htbp]
\centering
\begin{tabular}{lrrr}
\toprule
Size & Total (ns) & Bandwidth (GB/s) & Unidir.\ peak util.\ \\
\midrule
4 KiB   &        455{,}097 &   2.30 &  0.8\% \\
64 KiB  &        535{,}035 &  31.36 & 10.5\% \\
1 MiB   &      1{,}500{,}029 & 178.95 & 59.7\% \\
16 MiB  &     16{,}661{,}095 & 257.78 & \textbf{85.9\%} \\
\bottomrule
\end{tabular}
\caption{Multi-GPU K2K sustained bandwidth via NVLink P2P,
2$\times$ H100~NVL on Azure NC80adis\_H100\_v5 (NV12 topology:
12 NVLink4 links between the two GPUs, 25~GB/s per direction per
link).  Peak utilisation is reported against the 300~GB/s
unidirectional peak of the bundle (equivalently, 43.0\% of the
600~GB/s bidirectional peak at 16~MiB).  At 16~MiB the runtime
reaches 85.9\% of the unidirectional peak --- the stream is
saturating the fabric in the direction it is measured.  The
small-payload path is dominated by per-call overhead, not
bandwidth.  This is consistent with the cross-GPU K2K design's
expectation that the inter-GPU tier be bandwidth-bound, not
setup-bound, at the sizes where it matters operationally.}
\label{tab:mgpu-bw}
\end{table}

A complementary one-shot latency measurement at the same sizes
(\Cref{tab:migration} below) characterizes the small-payload
regime: NVLink P2P is 1.96$\times$ faster than host-staged at
1~KiB, growing to 8.65$\times$ at 16~MiB.  The two tables
together cover the bandwidth/latency Pareto front for
cross-GPU actor migration.

\begin{table}[htbp]
\centering
\begin{tabular}{lrrr}
\toprule
Size   & NVLink p99 (ns) & Host p99 (ns) & P2P speedup (mean) \\
\midrule
1 KiB   &      8{,}029 &     13{,}840 & 1.96$\times$ \\
4 KiB   &      7{,}850 &     16{,}770 & 2.02$\times$ \\
64 KiB  &      8{,}180 &     20{,}120 & 2.44$\times$ \\
1 MiB   &     12{,}630 &     57{,}510 & 5.08$\times$ \\
16 MiB  &     71{,}439 &    603{,}605 & \textbf{8.65$\times$} \\
\bottomrule
\end{tabular}
\caption{NVLink P2P vs.\ host-staged migration latency, 200 trials
per cell, 50 warmup iterations.  At small payloads the per-call
driver overhead dominates and NVLink wins by $\sim$2$\times$; at
16~MiB the bandwidth gap dominates and NVLink wins by 8.65$\times$.
The host path's per-MiB cost grows roughly linearly with size
because the data crosses the PCIe + host-memory boundary twice.}
\label{tab:migration}
\end{table}

\Cref{thm:fast-path-selection} of \Cref{sec:migration:cross-gpu}
guarantees by construction that NVLink-paired actors never take
the host path; \Cref{tab:migration} quantifies the associated
speedup, from $\sim$2$\times$ at the K2K scale to $\sim$9$\times$
at the bulk-state-transfer scale.

\subsection{The Cost of Actor Semantics: WaveSim3D and Application Workloads}
\label{sec:eval:apps}

A separate question from the control-path result of
\Cref{sec:eval:injection} is: once one pays for actor semantics---a
lifecycle state machine, per-actor supervision, a K2K mailbox, an
HLC, a tenant tag---how much throughput is given up on a workload
where none of those are strictly necessary?  We measure this
against a WaveSim3D stencil (3D FDTD, $64^3$), run twice: once as a
hand-tuned fused GPU stencil and once as a population of block
actors implementing the same update rule under the \system{}
runtime.  The difference is one measure of the abstraction cost.

\begin{table}[htbp]
\centering
\begin{tabular}{lrrr}
\toprule
Workload & Backend & Throughput & Speedup vs.\ CPU \\
\midrule
WaveSim3D (FDTD, $64^3$)        & GPU stencil    & 78{,}089 Mcell/s & 217.9$\times$ \\
WaveSim3D (FDTD, $64^3$)        & GPU block actor & 18{,}633 Mcell/s & 52.1$\times$ \\
TxMon (fraud detection)          & CUDA codegen   & 205.31 G~tx/s & 15{,}624$\times$ \\
ProcInt (process mining DFG)     & GPU            & 10.83 M~ev/s & --- \\
\bottomrule
\end{tabular}
\caption{Three application workloads on H100~NVL.  Speedups are
relative to CPU baseline (Rayon, 40 cores).  WaveSim3D is reported
with two backends to quantify the cost of the actor abstraction
over a hand-tuned stencil kernel: a 4.2$\times$ slowdown for the
actor model, which is the price of the lifecycle, mailbox, and
supervision machinery on a workload that does not need them.}
\label{tab:apps}
\end{table}

The WaveSim3D comparison isolates the \emph{price of actor
semantics} on a workload that has no structural use for most of
them.  4.2$\times$ is not small; we report it in the body of the
evaluation rather than behind a disclaimer.  The reading we would
encourage is that a purely-numerical workload with a fused stencil
kernel should remain a fused stencil kernel; the actor abstraction
tends to pay off only on workloads where at least one of
lifecycle management, supervisor restart, cross-actor messaging,
cross-tenant isolation, or cross-GPU migration is a required
feature rather than an incidental one.  TxMon's fraud-detection
pipeline and ProcInt's process-mining DFG fall into that category
(both use per-tenant routing and supervisor restart); WaveSim3D
does not.  The actor framework's cost is most plausibly amortized
across applications that need its primitives.

\subsection{Mechanized Verification Coverage}
\label{sec:eval:tlc}

The four core theorems of \Cref{thm:lifecycle-soundness,%
thm:no-message-loss,thm:hlc-restart-monotone,thm:cross-tenant-isolation}
plus the migration-safety and fast-path-selection theorems of
\Cref{sec:migration} are mechanized in six \TLA{} modules under
\texttt{docs/verification/}.  \Cref{tab:tlc-stats} reports the
state-space size and wall-clock for each spec under its bounded
configuration; all six pass with no counterexamples within their
stated bounds.  An earlier iteration of the migration spec did not
pass: TLC produced a counterexample on the \texttt{ChecksumMatch}
invariant when late-arriving messages mutated the source's state
variable between capture and swap.  The counterexample identified a
specification-level bug—the spec did not model the immutability of
the captured snapshot that the Rust implementation already
enforced.  \Cref{sec:disc:tlc-bug} describes the episode in detail.
We mention it here because it is a concrete indication that the
verification apparatus is doing engineering work rather than
serving only as proof dressing.

\begin{table}[htbp]
\centering
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lrrrrl}
\toprule
Spec & Bounds & Distinct states & Generated & Wall (s) & Result \\
\midrule
\texttt{actor\_lifecycle} & $|A|{\le}3,$ Steps${\le}8$              &   2{,}681 &   7{,}015 & $<$1 & OK \\
\texttt{hlc}              & $|N|{\le}2,$ Phys${\le}3$               & 4{,}600{,}000 & 16{,}700{,}000 &   29 & OK \\
\texttt{k2k\_delivery}    & $|A|{\le}2,$ Msgs${\le}4,$ QCap${\le}2$ &     241 &     399 & $<$1 & OK \\
\texttt{tenant\_isolation}& $|T|{\le}2,$ $|K|{\le}3,$ Msgs${\le}3$ & 12{,}261 & 26{,}953 & $<$1 & OK \\
\texttt{migration}        & Msgs${\le}4$                            &     725 &   1{,}445 & $<$1 & OK \\
\texttt{multi\_gpu\_k2k}  & $|G|{\le}2,$ $|A|{\le}3,$ Msgs${\le}3$ &   3{,}452 &   9{,}571 & $<$1 & OK \\
\bottomrule
\end{tabular}
\caption{TLC model-checking statistics for the six \TLA{}
specifications.  ``Steps'', ``Msgs'', ``Phys'' and ``QCap'' are
the \texttt{MaxSteps}, \texttt{MaxMsgs}, \texttt{MaxPhys} and
\texttt{QueueCap} configuration parameters respectively, in the
notation of the specs' \texttt{.cfg} files.  Run on TLC 2.19
(Java 17, 8 worker threads on the 80-vCPU EPYC host).  All specs
pass under the bounded configurations listed; no invariant
violations were observed.  The \texttt{hlc} spec has the largest
state space because the Hybrid Logical Clock state vector is
structurally larger than the other modules' state.}
\label{tab:tlc-stats}
\end{table}

\subsection{Reproducibility}
\label{sec:eval:repro}

All measurements are reproducible from the open-source distribution
of \system{} at the commit recorded in
\texttt{docs/benchmarks/h100-b200-baseline.md} (commit 42724ae,
RingKernel 0.4.2; the equivalent v1.0 release artifact is on the
release branch).  The full benchmark suite is invokable as
\texttt{cargo bench --workspace -p ringkernel-cuda --features cuda},
which produces the Criterion HTML reports under
\texttt{target/criterion/}; the methodology document
\texttt{docs/benchmarks/METHODOLOGY.md} describes the GPU clock
locking, exclusive compute mode, and ECC monitoring protocol step
by step, and \texttt{docs/benchmarks/h100-b200-baseline.md} captures
the full system state at the moment of the reported measurements.
