% sections/implementation.tex — Implementation: \system{}
\section{Implementation: \system{}}
\label{sec:implementation}

\system{} is a Rust workspace with 22 crates totaling roughly 80,000
lines, of which the core
runtime—\texttt{ringkernel-core},
\texttt{ringkernel-cuda}, and
\texttt{ringkernel-cuda-codegen}—accounts for the formal model's
implementation surface.  The remaining crates implement domain
adapters (audio FFT, accelerated networking, transactional
monitoring) and language bindings (Python, CLI) that sit outside
this paper's scope.  We focus here on the three crates that realize
the formal model and discuss the engineering choices that made each
small-step rule expressible against CUDA's driver-API surface.

\subsection{Crate Structure}
\label{sec:implementation:crates}

\begin{itemize}
\item \texttt{ringkernel-core} (\(\sim\)18,000 LOC): host-side
runtime. Implements the supervisor, the K2K bus, the HLC, the
audit-tag registry, the dead-letter store, and the snapshot--restart
protocol. Contains the host-side state of every model component
(\(\Theta\), \(M\), \(I\)).

\item \texttt{ringkernel-cuda} (\(\sim\)9,000 LOC of Rust + 4,000
LOC of CUDA): device-side runtime. Hosts the persistent kernel, the
mapped-memory control block, the SMEM/DSMEM/HBM ring buffers, the
release-acquire fence sequences, and the per-actor tick loop.

\item \texttt{ringkernel-cuda-codegen} (\(\sim\)6,000 LOC): a
build-time transpiler that lifts user-supplied Rust actor bodies into
CUDA C++ that links against the persistent kernel. The lift is
syntax-directed; the formal model treats the resulting code as the
\(\delta\) function of \Cref{rule:process}.
\end{itemize}

\subsection{CUDA Driver-API Surface}
\label{sec:implementation:driver}

The runtime uses the CUDA driver API rather than the higher-level
runtime API for two reasons: precise control over stream and event
ordering (required for the snapshot atomicity argument of
\Cref{sec:memory:snapshot}), and access to the cooperative-launch
and cluster primitives (required for the persistent kernel itself).
The set of driver entry points used is:

\begin{itemize}
\item \texttt{cuLaunchCooperativeKernel}: launch the persistent
kernel with grid-wide synchronization enabled. Called once at
runtime startup.

\item \texttt{cuMemAllocAsync}, \texttt{cuMemFreeAsync}: stream-ordered
allocation. The actor's per-instance state (mailbox slots,
scratch buffers) is allocated this way so that allocation cost is
overlapped with the actor's normal execution rather than serialized
on a global mutex.

\item \texttt{cuMemHostAlloc} with the \texttt{CU\_MEMHOSTALLOC\_\\
DEVICEMAP} flag: pinned, mapped host memory. This is the substrate
for the control block (the host writes commands into mapped memory;
the device polls them at the start of each tick).  The host
amortizes command submission through the CPU's write-combining
buffer: a 1{,}000-command burst costs 0.18~$\mu$s wall-clock
(\Cref{tab:injection}).

\item \texttt{cuStreamWaitValue32} / \texttt{cuStreamWriteValue32}:
host-side polling on a device-resident 32-bit flag. Used by the
supervisor to await the kernel's acknowledgement of a control
command (e.g., quiesce-complete, snapshot-taken) without resorting
to event synchronization.

\item \texttt{cuEventRecord} / \texttt{cuEventSynchronize}: precise
ordering between supervisor-initiated host-to-device copies (snapshot
restoration, control block updates) and the persistent kernel's tick
loop. Used in the critical section of \textsc{Restart}.

\item \texttt{cuMemcpyAsync}: bulk host-to-device and device-to-device
copies for snapshot capture and restoration.
\end{itemize}

The full set of driver calls fits on a single page; the engineering
work in \texttt{ringkernel-cuda} is in the careful sequencing of
these calls under the constraints of the formal model rather than
in the breadth of API consumed.

\subsection{The Persistent Kernel}
\label{sec:implementation:kernel}

The kernel is a single CUDA C++ function launched once at startup
and never returning until shutdown.  Its outer structure is:

\begin{verbatim}
__global__ void persistent_kernel(ControlBlock* cb,
                                  ActorState*  states,
                                  K2KBus*      bus) {
  auto cluster = cooperative_groups::this_cluster();
  auto grid    = cooperative_groups::this_grid();
  while (!cb->shutdown) {
    process_control(cb);            // poll mapped memory
    deliver_phase(states, bus);     // K2K Deliver, per-actor
    process_phase(states, bus);     // Process step, per-actor
    cluster.sync();                 // intra-cluster barrier
    bus->forward_phase(cluster);    // DSMEM cross-block hop
    grid.sync();                    // grid-wide barrier
  }
}
\end{verbatim}

The two synchronization barriers per iteration (\texttt{cluster.sync()}
and \texttt{grid.sync()}) realize the BSP (bulk-synchronous parallel)
discipline: every actor advances at most one tick per iteration,
and no actor observes a message produced in the same iteration as
its own \textsc{Process} step.  This is the operational model that
the formal calculus's small-step semantics abstracts; the formal
``one rule per step'' becomes ``one tick per iteration of the
persistent kernel.''

The control block is the host's only handle on the running kernel.
A 64-byte cache-aligned struct in mapped memory, it carries:
the shutdown flag; per-actor command slots (start, stop, restart,
snapshot); the supervisor's restart counters; and a 32-bit
acknowledgement field that the kernel writes after each tick.  The
host writes commands by ordinary memory store; the device sees them
on the next tick's \texttt{process\_control} call.  No CUDA driver
call is involved in the steady state.

\subsection{K2K Bus Implementation}
\label{sec:implementation:k2k}

The K2K bus realizes the calculus of \Cref{sec:k2k} through three
ring-buffer implementations, one per channel tier, all sharing the
same envelope layout (\Cref{def:k2k-envelope}).

\paragraph{SMEM ring}
A 4~KiB ring inside each block's shared memory; producer and
consumer cursors are 32-bit \texttt{atomicAdd} counters. Used only
when the destination actor lives in the same block.  Latency is
sub-cycle (the cursor update overlaps with the payload write).

\paragraph{DSMEM ring}
A 16~KiB ring split across cluster blocks via
\texttt{cluster.map\_shared\_rank()}. The ring is partitioned so that
each block owns the ring slots into which it writes; the consumer
reads via the cluster-shared pointer. The release-acquire pair from
\Cref{sec:memory:consistency} is realized as a cluster-scoped
\texttt{cuda::atomic\_thread\_fence} in the producer and a matching
fence in the consumer.

\paragraph{HBM ring}
A 1~MiB ring in HBM per (source-cluster, destination-cluster) pair.
The producer enqueues with a global-scoped
\texttt{cuda::atomic\_ref<...>::fetch\_add(release)}; the consumer
dequeues with a matching \texttt{load(acquire)}.  This is the
slowest path ($\sim$400 cycles round trip) and is used only when
the destination is in a different cluster.

The dead-letter store is a single per-tag HBM ring with capacity
configurable from the host (default 16,384 envelopes); the
idempotency cache is a Cuckoo hash with two tables, total capacity
\(2^{16}\) entries per tag, evicting on insert collision.  Both
data structures are entirely device-resident in steady state; the
host inspects them on demand via a debug entry point.

\subsection{Snapshot-Restart Engineering}
\label{sec:implementation:snapshot}

\Cref{sec:supervision:restart}'s three-step protocol (capture,
reset, reactivate) is realized in
\texttt{ringkernel-\allowbreak core/\allowbreak
src/supervisor.rs} as a single
\texttt{Mutex}-guarded host critical section.  The three steps
are:

\begin{enumerate}
\item \emph{Capture}: write the actor's HLC into a per-actor
\texttt{ret\_hlc} field in the control block; await the kernel's
ack via \texttt{cuStreamWaitValue32}.

\item \emph{Reset}: issue a \texttt{cuMemcpyAsync} from the actor's
\(\mathit{snap}\) HBM region to its \(\mathit{live}\) HBM region;
record a CUDA event; await it via
\texttt{cuEventSynchronize}.

\item \emph{Reactivate}: write the lifecycle-state flag in the
control block from \Failed{} to \Initializing{}; the kernel's next
tick observes this and runs the actor's per-actor reactivation
code, which (per \Cref{rule:activate}) takes a fresh snapshot equal
to the just-restored \(\mathit{live}\).
\end{enumerate}

The mutex serializes restart against any concurrent supervisor
action on the same actor (e.g., a parallel attempt to issue a
\textsc{Quiesce} or to take an out-of-band snapshot). The cost of
the full protocol is dominated by the event-synchronization
round-trip of step~3, not by the device-side copy of step~2.  We
measured this directly on H100~NVL (200 trials per state size,
\Cref{tab:snapshot}); a 4~KiB actor state restores in
10.6~$\mu$s, of which 7.3~$\mu$s is the event sync and only
1.9~$\mu$s is the actual copy.  The protocol scales to 64~KiB at
the same total cost (the copy is small enough that HBM bandwidth
is irrelevant).

\begin{table}[htbp]
\centering
\begin{tabular}{lrrrr}
\toprule
State size & capture (ns) & copy (ns) & ack (ns) & total (ns) \\
\midrule
1 KiB  & 1{,}331 & 1{,}893 & 7{,}032 & 10{,}256 \\
4 KiB  & 1{,}319 & 1{,}913 & 7{,}264 & 10{,}496 \\
16 KiB & 1{,}300 & 1{,}980 & 7{,}269 & 10{,}549 \\
64 KiB & 1{,}300 & 1{,}980 & 7{,}300 & 10{,}580 \\
\bottomrule
\end{tabular}
\caption{Snapshot--restart per-phase wall-clock, 200 trials per
size.  Total is dominated by the event-synchronization ack; the
device-side copy contribution is sub-2~$\mu$s even at 64~KiB.}
\label{tab:snapshot}
\end{table}

\subsection{Admin Surface and RBAC}
\label{sec:implementation:admin}

The supervisor exposes its operations to the host via an HTTP/JSON
admin API in \texttt{ringkernel-\allowbreak core/\allowbreak
src/admin.rs}.  The endpoints relevant to the formal model are:

\begin{itemize}
\item \texttt{POST\ /api/\allowbreak v1/\allowbreak admin/\allowbreak
k2k/\allowbreak dlq/\allowbreak replay}: replay a dead-letter
envelope per \textsc{Replay}~(\Cref{rule:k2k-replay}).
\item \texttt{GET\ /api/\allowbreak v1/\allowbreak admin/\allowbreak
k2k/\allowbreak dlq/\allowbreak list?tag=...}: list
dead-letters for a given audit tag.
\item \texttt{POST\ /api/\allowbreak v1/\allowbreak
admin/\allowbreak actor/\allowbreak restart}: force a
\textsc{Restart}~(\Cref{rule:restart}) on a named actor.
\item \texttt{GET\ /api/\allowbreak v1/\allowbreak admin/\allowbreak
actor/:id/\allowbreak snapshot}: download the
\(\mathit{snap}\) bytes for offline inspection.
\item \texttt{POST\ /api/\allowbreak v1/\allowbreak
admin/\allowbreak config}: hot-reload the restart-policy and
backpressure configuration without restarting the runtime.
\end{itemize}

The admin API is gated by an RBAC layer (capability-based; the
identity carries a set of \texttt{can\_replay},
\texttt{can\_restart}, \texttt{can\_inspect} capabilities) so that
the operational levers are not casually accessible.  The audit tag
of every admin invocation is recorded in the audit trace
(\Cref{sec:tenancy:audit}); a replay of an envelope with audit tag
\(\tau\) is itself tagged \(\tau\).

\subsection{Test Suite}
\label{sec:implementation:tests}

\system{} ships with 1{,}496 unit and integration tests under
\texttt{cargo test --workspace}.  The tests fall into four
categories:

\begin{enumerate}
\item \emph{Unit tests of host-side data structures}
($\sim$900 tests): the supervisor, the K2K queues, the dead-letter
store, the idempotency cache, the HLC, the audit-tag registry.
These run in $\sim$15~s on the dev box.

\item \emph{Integration tests of the runtime} ($\sim$400 tests):
end-to-end actor lifecycles, cross-tenant rejection, snapshot
restoration, restart policies, replay flow.  These run in
$\sim$45~s.

\item \emph{Device-side smoke tests} (74 tests, gated on
\texttt{--features cuda}): the persistent kernel, the K2K rings, the
DSMEM cross-block hop, the cooperative-launch path. These require a
GPU and run in $\sim$60~s on H100.

\item \emph{Property-based tests} (122 tests using
\texttt{proptest}): randomized exploration of the actor lifecycle
and K2K calculus, checking the invariants of
\Cref{thm:lifecycle-soundness,thm:no-message-loss,thm:fifo-per-sender,%
thm:cross-tenant-isolation} on randomly-generated traces.
\end{enumerate}

The property-based tests are an empirical complement to the bounded
\TLA{} verification: where TLC exhausts state space at a given
configuration, \texttt{proptest} samples randomly from a much
larger configuration space.  The combination has caught
implementation regressions (e.g., a concurrency bug in the
dead-letter store evicted on the wrong cursor) that escaped both
the unit tests and the bounded TLA model.
