% sections/related_work.tex — Related Work
\section{Related Work}
\label{sec:related_work}

\system{} sits at the intersection of four research streams that
have, until now, largely evolved independently: actor calculi and
their type systems; actor frameworks extended to heterogeneous or
GPU execution; GPU programming abstractions, including device-side
and task-based runtimes; and mechanized formal methods for
concurrent and distributed systems—including a small but relevant
literature on verifying GPU kernels directly.  We review each in
turn, positioning \system{} relative to representative work.  The
short version of the positioning is: prior work has addressed
\emph{actor-to-accelerator integration}, \emph{formal reasoning
about GPU kernel semantics}, and \emph{device-side/task-based
GPU execution} as three largely separate threads.  \system{}'s
claim is not to improve on any of those individually; it is to
combine \emph{persistent device residency, actor lifecycle,
supervision, HLC, tenant routing, and cross-GPU migration} in a
single formal and runtime package, with each property stated as a
theorem over a small-step calculus and each theorem bounded-model-checked
in \TLA{}.

\subsection{Actor Calculi and Type Systems}
\label{sec:rw:actors}

The lineage from Hewitt's 1973 informal model~\citep{hewitt1973actors}
to Agha's transition-system formalization~\citep{agha1986actors}
established the operational semantics on which essentially all
subsequent work rests.  Practical actor systems—Erlang/OTP
\citep{armstrong2007erlang}, Akka~\citep{akka2010},
Pony~\citep{clebsch2015pony}—have layered type systems, capability
systems, and supervision disciplines on top of Agha's small-step
core, but the core itself has remained stable: an actor is a
mailbox-driven sequential process; messages are values; state is
private.

\system{}'s lifecycle calculus (\Cref{sec:model:lifecycle}) inherits
this core.  What differs is the material that surrounds it.  Where Agha's calculus
treats actor creation as a single-step operation that immediately
produces an active actor, our model interposes \Initializing{} and
\Active{} states to capture the snapshot-restoration cost of GPU
restart.  Where Pony's deny capabilities forbid concurrent reads
and writes through static type discipline~\citep{clebsch2015pony},
our cross-tenant isolation theorem (\Cref{thm:cross-tenant-isolation})
forbids cross-boundary message flow through structural routing—the
runtime obstacle is similar (you cannot reach the data you should not
see) but the mechanism is dynamic, not static.

\subsection{Actors Meet Accelerators}
\label{sec:rw:hetero-actors}

The closest prior work on integrating actor abstractions with
GPU-class hardware comes from the C++ Actor Framework (CAF).
Charousset et~al.~\citep{charousset2016manyfold} extended CAF to
heterogeneous many-core machines via OpenCL (\emph{Manyfold actors}),
and Hiesgen et~al.~\citep{hiesgen2018opencl} sharpened that work
into \emph{OpenCL Actors}, adding data parallelism to actor-based
programming with CAF as a first-class construct.  Both introduce
an actor whose \(\delta\) function is an OpenCL kernel spawned and
managed through the CAF runtime.  The emphasis is on CPU-side
orchestration: the actor's identity, mailbox, and lifecycle live
on the host, and the GPU is a device on which one particular
action of that actor runs.  \system{}'s move is the dual one: the
actor's identity, mailbox, lifecycle, supervisor, HLC, and tenant
tag all live \emph{on the device}, and the host issues commands
but never owns actor state.  None of the CAF-based heterogeneous
actor works state formal properties about device-resident
lifecycle, HLC monotonicity under restart, or cross-tenant
routing—those questions do not arise when the actor is
host-resident.

\subsection{GPU Programming Abstractions}
\label{sec:rw:gpu}

The persistent-kernel literature
\citep{aila2009raytrace,gupta2012persistent,steinberger2014whippletree}
is the closest GPU-side prior work; we discussed it in
\Cref{sec:bg:persistent}.  Four other strands deserve mention.

\paragraph{Domain-specific GPU compilers}
Triton~\citep{tillet2019triton}, JAX~\citep{jax2018}, and
TVM~\citep{chen2018tvm} all generate GPU kernels from higher-level
descriptions.  Their abstractions are functional (JAX, TVM) or
block-functional (Triton): a kernel is a function applied to
tensors, not a long-lived process.  None of them exposes a notion
of actor lifecycle, message passing, or causal time.  They are
complementary rather than competing: one could imagine a
Triton-generated kernel running inside a \system{} actor as the
actor's \textsc{Process} step.

\paragraph{Device-side and task-based GPU runtimes}
Two recent works address the host-orchestration overhead we attack
in \Cref{sec:eval:injection}, from different directions.
\emph{Mustard}~\citep{mustard2025} is a device-side execution model
for multi-GPU task graphs in which the CPU initializes the graph
once and the GPU thereafter schedules graph vertices, tracks
dependencies, and balances across devices without host
intervention.  \emph{Cypress}~\citep{cypress2025} is a task-based
programming model with sequential semantics targeting modern
(Hopper-class) GPUs, where the compiler infers parallelism and
inserts data movement under a sequential-execution abstraction.
Both eliminate per-step host orchestration, as \system{} does, but
their unit of abstraction is a task/task-graph vertex rather than a
long-lived actor: neither models lifecycle states, supervisor
restart, HLC, tenant routing, or migration.  The mechanisms are
complementary—\system{}'s \textsc{Process} step could in principle
delegate to a Cypress task or a Mustard sub-graph—but the formal
claim is orthogonal.

\paragraph{CUDA Graphs and stream-based composition}
CUDA Graphs~\citep{nvidia2023cooperative} let an application capture
a sequence of kernel launches and replay them as a single submission;
the speedup over individual launches is roughly 3$\times$ on H100
(per our measurements; see \Cref{sec:evaluation}).  CUDA Graphs
attack the same launch-overhead problem as \system{}, but at a
different layer: a CUDA Graph is still a sequence of one-shot
kernels, and the host still drives every iteration.  \system{}'s
persistent kernel runs continuously; the host injects work into
mapped memory but does not orchestrate the kernel's tick loop.

\paragraph{Comparison summary}
\Cref{tab:comparison} summarizes the design space along the axes
that the formal model isolates: programming-model granularity,
lifecycle abstraction, message-passing primitive, fault tolerance,
causal time, and multi-tenant isolation.  The table is intended as
a positioning aid rather than a performance comparison; for each
row, ``\cmark'' means the system provides a first-class abstraction
of the property, ``\textit{ext}'' means the property is achievable
by the user's own glue code, and ``--'' means there is no support.

\begin{table}[htbp]
\centering
\footnotesize
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.15}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
\toprule
System & Granularity & Lifecycle & Msg.\ pass & Fault tol. & Causal time & Multi-tenant \\
\midrule
\system{}
                            & actor      & \cmark   & \cmark & \cmark      & HLC      & \cmark \\
Persistent kernels$^{a}$
                            & loop body  & --       & queue  & --          & --       & -- \\
Cypress \citep{cypress2025}
                            & task       & --       & --     & --          & --       & -- \\
Mustard \citep{mustard2025}
                            & task graph & --       & --     & --          & --       & -- \\
CUDA Graphs                 & kernel     & --       & --     & --          & --       & -- \\
Triton / JAX / TVM$^{b}$
                            & function   & --       & --     & --          & --       & -- \\
OpenCL Actors \citep{hiesgen2018opencl}
                            & host actor & \cmark   & \cmark & \textit{ext}& seq.\ ID & -- \\
Erlang OTP \citep{armstrong2007erlang}
                            & actor      & \cmark   & \cmark & \cmark      & seq.\ ID & \textit{ext} \\
Akka \citep{akka2010}       & actor      & \cmark   & \cmark & \cmark      & seq.\ ID & \textit{ext} \\
Pony \citep{clebsch2015pony}
                            & actor      & \cmark   & \cmark & --          & --       & static \\
\bottomrule
\end{tabular}%
}
\vspace{2pt}
\\[-1pt]
{\footnotesize
\textsuperscript{a}\,\citet{aila2009raytrace,gupta2012persistent,%
steinberger2014whippletree}.\quad
\textsuperscript{b}\,\citet{tillet2019triton,jax2018,chen2018tvm}.}
\caption{Design-space comparison of \system{} against representative
GPU programming abstractions, heterogeneous actor frameworks, and
CPU actor systems.  ``\cmark'' = first-class abstraction;
``\textit{ext}'' = achievable via external glue;
``static'' = compile-time enforcement only;
``--'' = unsupported; ``HLC'' = Hybrid Logical Clock; ``seq.\ ID''
= monotone sequence number per actor.  GPU-native rows are above
the dividing line; heterogeneous and CPU-side actor systems are
shown below for context.  Among the systems summarized, none
packages lifecycle, causal clocks, tenant isolation, and migration
as a single first-class substrate running on the device.}
\label{tab:comparison}
\end{table}

The table's rightmost columns—causal time, multi-tenant isolation,
fault tolerance, lifecycle—are the formal-model territory.  The
GPU-side abstractions treat them as out-of-scope (CUDA Graphs,
Triton, JAX, TVM, Cypress, Mustard) or approximate them through a
single all-or-nothing pattern (the persistent-kernel literature).
Heterogeneous actor frameworks (OpenCL Actors, Manyfold Actors)
address lifecycle and messaging but keep actor identity on the
host.  The CPU-side actor systems handle lifecycle and supervision
at the operating-system layer that does not exist on the GPU.
\system{}'s contribution is a single device-resident substrate that
carries all four properties with a formal statement for each.

\paragraph{GPU memory consistency}
Lustig et~al.~\citep{lustig2014abstraction} and Hower
et~al.~\citep{hower2014} formalized the memory consistency models of
heterogeneous architectures, including GPUs.  Their models are
protocol-level rather than program-level: they describe what
orderings the hardware guarantees, not what programs can be written
on top.  Our memory-model section (\Cref{sec:memory:consistency})
adopts their framework as a parameter, constraining our
\textsc{SendOK} and \textsc{Deliver} rules to compose
release-acquire pairs that respect their axioms.  This is the
narrowest formal interface: we do not claim a new memory model;
we claim a runtime that uses an existing one correctly.

\subsection{Formal Methods for Concurrent and GPU Systems}
\label{sec:rw:formal}

The use of \TLA{}~\citep{lamport2002} for verifying distributed
systems, popularized by AWS~\citep{newcombe2014}, has produced a
methodology in which a target protocol is specified abstractly,
model-checked for invariants under bounded configurations, and then
mapped onto an implementation by an explicit correspondence
table.  We follow this methodology directly: each theorem in this
paper has a corresponding \TLA{} module in
\texttt{docs/verification/}, and each module's variables map onto
specific Rust modules in \texttt{crates/ringkernel-core/}.  The
correspondence table is in \texttt{docs/verification/README.md}.
The checks are bounded (small numbers of actors, messages, queue
capacities); we do not claim scale-free proofs.

\paragraph{Formal verification of GPU kernels}
A small but adjacent literature verifies the kernel programs
themselves rather than the surrounding runtime.  GPUVerify
\citep{betts2012gpuverify} introduced a synchronous,
delayed-visibility semantics for GPU programs and uses it to prove
data-race and barrier-divergence freedom of individual CUDA/OpenCL
kernels.  This is a different target from ours: GPUVerify reasons
about the internals of a single kernel, whereas \system{} reasons
about a long-lived persistent kernel hosting a population of
actors, with the small-step rules firing on message arrival,
supervisor event, and migration phase transition rather than on
intra-kernel thread scheduling.  The two lines of work are
compatible in principle—one could verify the persistent kernel's
internal synchronization with a GPUVerify-style tool while
verifying the actor calculus over the kernel at the \TLA{} level.

\paragraph{Causal time and HLC}
Logical clocks (\citet{lamport1978clocks}) and the more recent
hybrid construction of \citet{kulkarni2014hlc} are widely used in
distributed databases (MongoDB, CockroachDB) for cross-shard
causal consistency.  Our use is unusual in two ways: first, we apply
HLC \emph{within} a single device, not across a network; second,
we extend the standard monotonicity argument to cover supervisor-driven
restart, which has no analogue in the distributed-database setting
(databases use replication, not restart).  To our knowledge,
\Cref{thm:hlc-restart-monotone} is the first formal statement of
HLC monotonicity across an actor restart.

\paragraph{Information flow and tenant isolation}
The information-flow tradition stemming from
\citet{sabelfeld2003language} provides type systems and program
logics for proving non-interference between security levels.  Our
isolation theorem (\Cref{thm:cross-tenant-isolation}) is weaker than
non-interference in the static-type sense (we do not preclude
tenants from observing each other's resource usage through timing
or queue-depth side channels), but it is at the right granularity
for audit: an auditor's question is not ``did tenant~A learn
anything about tenant~B?''\ in the noninterference sense, but
``did any artifact of tenant~A appear in tenant~B's
delivery set?''\ which is exactly what our theorem rules out.

\paragraph{Reliable broadcast and at-least-once messaging}
The distributed-systems literature on reliable broadcast
(\citet{birman1987process}) and primary-backup
(\citet{junqueira2011zab}) has long-since established the
at-least-once / exactly-once / per-source-FIFO landscape for
message delivery.  Our K2K theorems
(\Cref{thm:no-message-loss,thm:fifo-per-sender,%
thm:no-duplicate-delivery}) are the within-device specializations
of those properties, with the additional twist that the dead-letter
discipline lets us state at-most-once observable delivery as a
conjunctive property over \(D\) and \(X\), not just \(D\).  The
\TLA{} structure is similar to Junqueira et~al.'s Zab
specification~\citep{junqueira2011zab}; what differs is the
within-device setting.
